---
title: "Reproducible Research - Week2, Course Project 1"
author: "Kerstin Funke"
date: "8/8/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---


## Loading and preprocessing the data

First, we will load the data, format the Date variable and build a new data frame which aggregates the number of steps per day.

```{r package, message=FALSE}
library(dplyr)
```

```{r activity}
activity <- read.csv("activity.csv", 
                     header = TRUE, 
                     sep = ",",
                     na.strings = "NA")

activity$date <- as.Date(activity$date, "%Y-%m-%d")

activityDay <- aggregate(activity$steps, 
                         by = list(date=activity$date),
                         FUN=sum)
activityDay<- rename(activityDay, steps = x)
```




## Mean total number of steps taken per day

Here you can see a histogram of steps taken per day.

```{r histogram}
hist(activityDay$steps, 
     xlab="Steps per Day", 
     main="Histogram of Steps per Day")
```

Here are the mean and the median of steps taken per day.

```{r mean and median}
mean(activityDay$steps,na.rm = TRUE)
median(activityDay$steps,na.rm = TRUE)
```

So the mean of the total number of steps taken per day is 10766.19 and the median is 10765.

## Average daily activity pattern

First, we build a new data frame which shows the average number of steps per interval across all days. 

```{r aggregate day}
activityInterval <- aggregate(activity$steps, 
                              by = list(interval=activity$interval), 
                              FUN=mean, na.rm = TRUE)
activityInterval<- rename(activityInterval, steps = x)
```

Then, we build a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r plot}
plot(activityInterval$interval, activityInterval$steps, 
     type="l", 
     xlab="5-minute interval", 
     ylab="average number of steps taken, averaged across all days", 
     main="Average daily activity pattern")
```

We then identify which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps. 

```{r max}
activityInterval$interval [which.max(activityInterval$steps)]
```

This shows that on average across all the days in the dataset the 5-minute interval at 8:35am shows the highest number of steps.



## Imputing missing values

First we determine the number of missing values in the original data set

```{r NA}
sum(!complete.cases(activity$steps))
```

Overall there are 2304 missing values. 

In a second step we develop a strategy to replace missing values: 
1) identifying the missing values
2) replacing missing values with the overall average per interval

We then set up a new data frame applying this strategy

```{r replace NA}
activityMissing <- activity
activityMissing$steps[is.na(activityMissing$steps)] <- mean(activityMissing$steps, na.rm=TRUE) 
```


As a next step we build a new data frame which aggregates the number of steps per day using the new data set 

```{r aggregate interval}
activityDayMissing <- aggregate(activityMissing$steps, 
                                by = list(date=activityMissing$date), 
                                FUN=sum)
activityDayMissing<- rename(activityDayMissing, steps = x)
```

Then we make a histogram of the total number of steps taken each day using this new data frame.

```{r histogram NA}
hist(activityDayMissing$steps, 
     xlab="Steps per Day", 
     main="Histogram of Steps per Day - replaced missing values")
```

We then recalculate the mean and median 

```{r mean and median NA}
mean(activityDayMissing$steps,na.rm = TRUE)
median(activityDayMissing$steps,na.rm = TRUE)
```

The results show that the mean stayed the same and the median now equals the mean. 



## Differences in activity patterns between weekdays and weekends

In order to identify potential differences in activity patterns between weekdays and weekends we set up new variables - first, a variable that identifies the different weekdays and then we recode those weekdays into a variable called 'weekend' that distinguishes between weekend and weekday.

```{r Weekend}
activityMissing$weekdays <- weekdays(activityMissing$date, abbreviate = FALSE)

activityMissing$weekend<-recode( activityMissing$weekdays, 
                                 "Saturday" = "Weekend", 
                                 "Sunday" = "Weekend", 
                                 .default = "Weekday")
```

Now we build a data set that shows how many steps participants took on average per interval on weekdays vs. weekends.

```{r aggregate interval missing values included}
activityDayMissingInterval <- aggregate(activityMissing$steps, 
                                       by = list(interval=activityMissing$interval, weekend=activityMissing$weekend), 
                                       FUN=mean, na.rm = TRUE)
activityDayMissingInterval <- rename(activityDayMissingInterval , steps = x)
```

In a second step we then set up two time series plots looking at the development throughout the day (using 5-minute intervals) on weekends (plot 1) and weekdays (plot 2)

```{r load lattice,  message=FALSE}
library(lattice)
```

```{r plot weekend weekday}
xyplot( steps ~ interval| weekend, 
        data =  activityDayMissingInterval, 
        type = "l", 
        layout=c(1,2), 
        ylab="Number of Steps", 
        xlab="Interval")
```